Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model

### 2021-10-22 analysis/paper/paper.Rmd
fix colours for Figure 3D

---
title: "Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model"
author: |
 Bjørn Peare Bartholdy$^1$, Amanda G. Henry$^1$
date: |
 $^1$ Department of Archaeological Sciences, Faculty of Archaeology, Leiden University, the Netherlands.
output:
  bookdown::pdf_document2:
    includes:
      in_header: "../templates/latex/pre-print/preamble.tex"
      #before_body: "../templates/latex/pre-print/before_body.tex"
    df_print: kable
    toc: no
    latex_engine: xelatex
    fig_caption: yes
    number_sections: yes
    keep_tex: true
    #extra_dependencies: "subfig"
  bookdown::html_document2:
    df_print: kable
    keep_md: true
  bookdown::word_document2:
    reference_docx: "../templates/template.docx"
#linestretch: 2.0 # double-spacing for review purposes
bibliography: refs.bib
csl: apa.csl
abstract: |
 Dental calculus has proven to contain a wealth of information 
 on the dietary habits of past populations. These insights have, to a large extent,
 been obtained by the extraction and identification of
 starch granules contained within the mineralised dental plaque from a wide range 
 of regions and time periods. The scope of previous studies have been limited to 
 microfossil extraction and identification to reconstruct dietary preferences from
 the archaeological record, and few studies have attempted to address the biases 
 of starch retention in dental calculus. Those that have considered this problem 
 have been limited to *in vivo* studies on modern humans and non-human primates.
 Here, we present a multispecies oral biofilm model, which allows experimental 
 research on  starch incorporation and retention to be conducted on *in vitro* 
 dental calculus in a controlled laboratory setting. The biofilms were exposed 
 to treatment solutions with known quantities of
 dietary starches (wheat and potato) during the 25-day growth period. After this,
 the starch granules were extracted from the mature biofilm (by dissolution in EDTA),
 and counted. We show that the granule counts extracted from the model dental calculus 
 represented a low proportion (ranging from 0.064% to 0.161%) of the total number 
 of granules exposed to the biofilms throughout the experiment.
 Additionally, we found that the ratios of granule sizes from the extracted 
 starch granules differed from the original treatment solutions, with large 
 granules (>20 $\mu$m) consistently being under-represented.
 We also found a correlation
 between the absolute granule counts and dry-weight of the biofilm 
 (*r* = 0.66, 90%CI[0.46,0.79]), as well
 as between the concentration (count per mg) of granules and dry-weight 
 (*r* = 0.30, 90%CI[0.06,0.51]).  
 Our results reinforce previous *in vivo* studies suggesting that dental calculus 
 presents a very small, and partly biased picture of the original dietary intake 
 of starches, with an over-representation of plants producing granules smaller than
 20 $\mu$m in size. The experimental model presented here is well-suited to address
 the need for further validation of methods and biases associated with dietary
 research on dental calculus.
keywords: 
  - oral biofilm
  - starch
  - alpha-amylase
  - dental calculus
---

>>>>

---
title: "Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model"
author: |
 Bjørn Peare Bartholdy$^1$, Amanda G. Henry$^1$
date: |
 $^1$ Department of Archaeological Sciences, Faculty of Archaeology, Leiden University, the Netherlands.
output:
  bookdown::pdf_document2:
    includes:
      in_header: "../templates/latex/pre-print/preamble.tex"
      #before_body: "../templates/latex/pre-print/before_body.tex"
    df_print: kable
    toc: no
    latex_engine: xelatex
    fig_caption: yes
    number_sections: yes
    keep_tex: true
    #extra_dependencies: "subfig"
  bookdown::html_document2:
    df_print: kable
    keep_md: true
  bookdown::word_document2:
    reference_docx: "../templates/template.docx"
#linestretch: 2.0 # double-spacing for review purposes
bibliography: refs.bib
csl: apa.csl
abstract: |
 Dental calculus has proven to contain a wealth of information 
 on the dietary habits of past populations. These insights have, to a large extent,
 been obtained by the extraction and identification of
 starch granules contained within the mineralised dental plaque from a wide range 
 of regions and time periods. The scope of previous studies have been limited to 
 microfossil extraction and identification to reconstruct dietary preferences from
 the archaeological record, and few studies have attempted to address the biases 
 of starch retention in dental calculus. Those that have considered this problem 
 have been limited to *in vivo* studies on modern humans and non-human primates.
 Here, we present a multispecies oral biofilm model, which allows experimental 
 research on  starch incorporation and retention to be conducted on *in vitro* 
 dental calculus in a controlled laboratory setting. The biofilms were exposed 
 to treatment solutions with known quantities of
 dietary starches (wheat and potato) during the 25-day growth period. After this,
 the starch granules were extracted from the mature biofilm (by dissolution in EDTA),
 and counted. We show that the granule counts extracted from the model dental calculus 
 represented a low proportion (ranging from 0.06% to 0.16%) of the total number 
 of granules exposed to the biofilms throughout the experiment.
 Additionally, we found that the ratios of granule sizes from the extracted 
 starch granules differed from the original treatment solutions, with large 
 granules (>20 $\mu$m) consistently being under-represented.
 We also found a correlation
 between the absolute granule counts and dry-weight of the biofilm 
 (*r* = 0.66, 90%CI[0.46,0.79]), as well
 as between the concentration (count per mg) of granules and dry-weight 
 (*r* = 0.30, 90%CI[0.06,0.51]).  
 Our results reinforce previous *in vivo* studies suggesting that dental calculus 
 presents a very small, and partly biased picture of the original dietary intake 
 of starches, with an over-representation of plants producing granules smaller than
 20 $\mu$m in size. The experimental model presented here is well-suited to address
 the need for further validation of methods and biases associated with dietary
 research on dental calculus.
keywords: 
  - oral biofilm
  - starch
  - alpha-amylase
  - dental calculus
---

### 2021-10-22 analysis/paper/paper.tex
fix colours for Figure 3D

\begin{document}
\maketitle
\begin{abstract}
Dental calculus has proven to contain a wealth of information
on the dietary habits of past populations. These insights have, to a large extent,
been obtained by the extraction and identification of
starch granules contained within the mineralised dental plaque from a wide range
of regions and time periods. The scope of previous studies have been limited to
microfossil extraction and identification to reconstruct dietary preferences from
the archaeological record, and few studies have attempted to address the biases
of starch retention in dental calculus. Those that have considered this problem
have been limited to \emph{in vivo} studies on modern humans and non-human primates.
Here, we present a multispecies oral biofilm model, which allows experimental
research on starch incorporation and retention to be conducted on \emph{in vitro}
dental calculus in a controlled laboratory setting. The biofilms were exposed
to treatment solutions with known quantities of
dietary starches (wheat and potato) during the 25-day growth period. After this,
the starch granules were extracted from the mature biofilm (by dissolution in EDTA),
and counted. We show that the granule counts extracted from the model dental calculus
represented a low proportion (ranging from 0.064\% to 0.161\%) of the total number
of granules exposed to the biofilms throughout the experiment.
Additionally, we found that the ratios of granule sizes from the extracted
starch granules differed from the original treatment solutions, with large
granules (\textgreater20 \(\mu\)m) consistently being under-represented.
We also found a correlation
between the absolute granule counts and dry-weight of the biofilm
(\emph{r} = 0.66, 90\%CI{[}0.46,0.79{]}), as well
as between the concentration (count per mg) of granules and dry-weight
(\emph{r} = 0.30, 90\%CI{[}0.06,0.51{]}).\\
Our results reinforce previous \emph{in vivo} studies suggesting that dental calculus
presents a very small, and partly biased picture of the original dietary intake
of starches, with an over-representation of plants producing granules smaller than
20 \(\mu\)m in size. The experimental model presented here is well-suited to address
the need for further validation of methods and biases associated with dietary
research on dental calculus.
\end{abstract}

>>>>

\begin{document}
\maketitle
\begin{abstract}
Dental calculus has proven to contain a wealth of information
on the dietary habits of past populations. These insights have, to a large extent,
been obtained by the extraction and identification of
starch granules contained within the mineralised dental plaque from a wide range
of regions and time periods. The scope of previous studies have been limited to
microfossil extraction and identification to reconstruct dietary preferences from
the archaeological record, and few studies have attempted to address the biases
of starch retention in dental calculus. Those that have considered this problem
have been limited to \emph{in vivo} studies on modern humans and non-human primates.
Here, we present a multispecies oral biofilm model, which allows experimental
research on starch incorporation and retention to be conducted on \emph{in vitro}
dental calculus in a controlled laboratory setting. The biofilms were exposed
to treatment solutions with known quantities of
dietary starches (wheat and potato) during the 25-day growth period. After this,
the starch granules were extracted from the mature biofilm (by dissolution in EDTA),
and counted. We show that the granule counts extracted from the model dental calculus
represented a low proportion (ranging from 0.06\% to 0.16\%) of the total number
of granules exposed to the biofilms throughout the experiment.
Additionally, we found that the ratios of granule sizes from the extracted
starch granules differed from the original treatment solutions, with large
granules (\textgreater20 \(\mu\)m) consistently being under-represented.
We also found a correlation
between the absolute granule counts and dry-weight of the biofilm
(\emph{r} = 0.66, 90\%CI{[}0.46,0.79{]}), as well
as between the concentration (count per mg) of granules and dry-weight
(\emph{r} = 0.30, 90\%CI{[}0.06,0.51{]}).\\
Our results reinforce previous \emph{in vivo} studies suggesting that dental calculus
presents a very small, and partly biased picture of the original dietary intake
of starches, with an over-representation of plants producing granules smaller than
20 \(\mu\)m in size. The experimental model presented here is well-suited to address
the need for further validation of methods and biases associated with dietary
research on dental calculus.
\end{abstract}
